---
title: Baxter microbiome example
author: Gavin Simpson (Based on Greenacre)
date: April 29, 2026
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup-options, echo = FALSE, results = "hide", message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(cache = TRUE, dev = 'svg', echo = TRUE, message = FALSE, warning = FALSE, fig.height=6, fig.width = 1.777777*6)
```

In this example, we will use a data set from Baxter *et al* (2016; *Genome Medicine*), who investigated the use of non-invasive tests for colorectal cancer. They studied the microbiomes of a number of patients and investigated whether there were detectable differences in microbial composition in stool samples between patients with cancer, versus adenomas, versus healthy individuals.

Load the packages we will use and the OTU data:

```{r load-baxter}
# load vegan, dplyr
library("vegan")
library("dplyr")

# load the data
baxter <- read.table(
  "https://raw.githubusercontent.com/michaelgreenacre/CODAinPractice/master/Baxter_OTU_table.txt",
  header = TRUE
  )
```

We need to remove the first three columns of the data as provided:

```{r}
baxter <- baxter |>
  as_tibble() |>
  select(-c(label:numOtus))
```

There are 490 samples and 335 OTUs in the data:

```{r}
dim(baxter)
```

The associated meta data are loaded using

```{r}
meta <- read.table(
  "https://raw.githubusercontent.com/michaelgreenacre/CODAinPractice/master/Baxter_Metadata.txt",
  header = TRUE
) |> as_tibble()

dim(meta)
names(meta)
meta
```

The data represent one of three types of cell (variable `dx`):

* `normal`,
* `cancer`, and
* `adenoma` (benign)

We need to convert ths variable to a factor so it can be used in the ordination method. We also centre and standardize the `Age` variable

```{r}
meta <- meta |>
  mutate(
    dx = factor(dx),
    std_Age = (Age - mean(Age)) / sd(Age)
  )
```

The species data are counts of the OTUs. We will look at two ordination methods in this brief example:

* CCA on the fourth-root transformed percentage abundances, and

* RDA on the CLR-transformed percentage abundances.

## CCA

We begin by converting the data to proportions

```{r}
baxter_prop <- baxter |> decostand(method = "total")
```

and then to stabilise the variance of the data we could use a square root transform, but we follow Greenacre (2021; *Annual Review of Statistics and its Applications*) and use a quarter root transformation

```{r}
baxter_qroot <- baxter_prop^0.25
```

The aim of this brief example is to explore the effects of `dx` (diagnosis) and `std_Age` on the OTU composition. We can fit this model using

```{r}
baxter_cca <- cca(baxter_qroot ~ dx + std_Age, data = meta)
```

If we plot these data, we notice a considerable outlier, sample 371:

```{r}
plot(baxter_cca, display = "sites", type = "text")
```

We can delete this observation and refit the model

```{r}
baxter_qroot <- baxter_qroot[-371, ]
meta <- meta[-371, ]
baxter_cca <- cca(baxter_qroot ~ dx + std_Age, data = meta)
```

```{r}
plot(baxter_cca, display = "sites", type = "text")
```

The two covariates in the model explain a relatively small amount of the variation in the OTUs

```{r}
baxter_cca
```

but such small proportions of variation explained are not necessarily an issue; with such a large number of taxa, even 490 samples are very sparsely scattered around the 335 dimensional space represented by the OTUs.

We can use the `anova()` method to test the overall model
```{r}
set.seed(2026-4-29)
anova(baxter_cca, parallel = 4)
```
the individual axes of the CCA
```{r}
set.seed(2026-4-29)
anova(baxter_cca, by = "axis", parallel = 4)
```
and the marginal effects of the covariates
```{r}
set.seed(2026-4-29)
anova(baxter_cca, by = "margin", parallel = 4)
```

To visualise the effects of the covariates, we can use a couple of *vegan*'s plotting commands

```{r}
baxter_cca |> plot(
  display = c("sites", "bp", "cn"),
  scaling = "sites"
)
ordiellipse(
  baxter_cca, meta$dx,
  kind = "se", conf = 0.95,
  col = palette.colors(3),
  scaling = "sites", lwd = 2, label = TRUE
)
```

It's not an especially pretty plot, but we see that the first axis captures the contrast between cancerous and normal/adenoma samples, while the second axis reflects the age of the patient.

### CLR transformation

To look at the CLR transformation, we start with the original data, `baxter` and use `decostand()` to apply the transformation. We use a pseudocount of 1.

```{r}
pseudo <- min(baxter_prop[baxter_prop > 0])
baxter_clr <- baxter_prop |>
  decostand(method = "clr", pseudocount = pseudo)
baxter_clr <- baxter_clr[-371, ]
```

We fit the same model as we did with CCA:

```{r}
baxter_rda <- rda(baxter_clr ~ dx + std_Age, data = meta)
```

and apply the same tests; for brevity, I just show the results for the marginal effects:
```{r}
set.seed(2026-4-29)
anova(baxter_rda, by = "margin", parallel = 4)
```

A plot that is superficially the same as the one we achieved using CCA is obtained from the RDA results:
```{r}
baxter_rda |> plot(
  display = c("sites", "bp", "cn"),
  scaling = "sites"
)
ordiellipse(
  baxter_rda, meta$dx,
  kind = "se", conf = 0.95,
  col = palette.colors(3),
  scaling = "sites", lwd = 2, label = TRUE
)
```
